---
title: "FEMA National Risk Index (NRI)"
subtitle: "sub. **FIX LATER!**"
description: "Visualizing FEMA NRI Counties in California Compared To Other States6"
author: "Marie Tolteca"
date: 1/22/2025
toc: true
code-line-numbers: true
editor_options: 
  chunk_output_type: console
---

# Library Packages
```{r, message=FALSE}
library(tidyverse)
library(here)
```

# Read in Data
```{r}
nri <- read_csv(here("data","National_Risk_Index_Counties_807384124455672111.csv")) %>% 
  janitor::clean_names()
```
IV. Build your visualization
Create a data viz that helps to answer the question, How do FEMA National Risk Index scores for counties in California compare to those in other states? This may require some data wrangling.

BEFORE you start wrangling / building your viz, be sure to…
identify / jot down your variables of interest and consider which data types they are
use online tools like from Data to Viz to help determine appropriate graph types, given your variables
roughly sketch out your plots by hand (see why this step matters!)
```{r}
unique(nri$state_name)
unique(nri$state_name_abbreviation)

```



Your final visualization should:

include data for the 50 US states only (no territories)
include a title (short, descriptive) & subtitle (describes main takeaway) (see Fundamentals of Data Visualization, Ch 22 for an example), a caption (describes the data source, e.g. “Data: FEMA National Risk Index (2025 Release)”), and alt text (following the formula you recently practiced in lab; use the fig-alt code chunk option to apply your alt text)
consider and implement strategies for highlighting trends / important information (e.g. arranging data, highlighting data, adjusting scales)
use custom colors (if applicable), rather than ggplot defaults
have an updated / polished theme
